---
title: "Setup a Play-by-Play Database with nflfastR"
author: "Nolan MacDonald"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setup a Play-by-Play Database with nflfastR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references/references.bib
csl: references/ieee.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

```{r setup, echo = FALSE, message = FALSE, eval=FALSE}
library(DBI)
library(RSQLite)
library(dplyr)  # Load dplyr to use the pipe operator
library(nflfastR)  # Load your relevant packages
library(stringr)
library(nuclearff)
```

## Install Packages

Data from `nflfastR` can be stored in a database to access locally instead of pulling play-by-play information each time.
The database takes up a significant amount of memory, as it stores data dating back to the 1999 season.
Working with the database is beneficial as is allows you to only bring into memory the data you actually need.
Using R, it is relatively easy to work with databases as long as you are comfortable with `dplyr` functions to manipulate and tidy data.

To use the database, the `DBI` and `RSQLite` packages should be installed.
To install these required packages, use the following commands in the R console:
```bash
install.packages("DBI")
install.packages("RSQLite")
```

## Play-By-Play Database

Now that the packages have been installed, the database can be built.
`nflfastR` uses the function `update_db()` to work with databases.
Using `update_db()` with no arguments will build a database with default arguments, where the SQLite database will be stored as `pbp_db` that contains all of the play-by-play data in a table called `nflfastR_pbp`.
To store the database in a specified directory, use the argument `dbdir`.
To specify the file name use `dbname` and to change the table name use `tblname`.

If there is already an existing database and you want to rebuild it from scratch, use the argument `force_rebuild=TRUE`. 
This is typically performed when there is an update to play-by-play data from fixing a bug or you want to wipe the database and update it. 
If you want to rebuild only specific seasons, you can use the command `force_rebuild=2023` or for multiple seasons, `force_rebuild=c(2022, 2023)`.

**Summary of Database and Arguments:**

- `update_db()`: Build or update play-by-play database
  - `update_db(dbdir = "path/to/save/db")`: Build play-by-play database and save to path
  - `update_db(dbname="db_name")`: Build play-by-play database with custom name (Default: `nflfastR_pbp`)
  - `update_db(tblname="tblname")`: Build play-by-play database with custom table name (Default: `nflfastR_pbp`)
  - `update_db(force_rebuild=TRUE)`: Rebuild entire play-by-play database
  - - `update_db(force_rebuild=c(2022, 2023))`: Rebuild play-by-play database 2022-2023 seasons

## Create Database

To first create and dump a database you need to load the libraries that were recently installed.
After loading the libraries, `update_db()` is utilized to create or update the database.
In this example, the database is saved in the repository's directory, `data/pbp_db`.

```{r, eval=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
# Import packages
library(DBI)
library(RSQLite)

# Create or update the database
nflfastR::update_db(dbdir = "data/pbp_db")
```

## Load Database

The database is `pbp_db` loaded from the directory `data/pbp_db`.
Once connected to the database with `dbConnect()`, the tables stored in the
database are obtained with `dbListTables`.
There should only be one table with the newly constructed database, `nflfastR_pbp`,
as previously discussed since the default naming schemes were used.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Connect to database
connect <- DBI::dbConnect(RSQLite::SQLite(), "./data/pbp_db/pbp_db")

# List tables
tables <- DBI::dbListTables(connect)
# Close the database when finished
DBI::dbDisconnect(connect)

tables
```

Another option is to look at the fields (columns) that are stored in the table.
Use `dbListFields()` to show the data available in the table.
For the example below, the fields are saved in a list and `head()` is used to
show only the first 10 fields since there is a large amount of data (372 fields!).

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Connect to database
connect <- DBI::dbConnect(RSQLite::SQLite(), "./data/pbp_db/pbp_db")

# List fields
fields <- DBI::dbListFields(connect, "nflfastR_pbp")
# Close the database when finished
DBI::dbDisconnect(connect)

head(fields, 10)
```

Finally, the play-by-play table is loaded from the database with `dplyr::tbl()`.
Use the play-by-play information to write code and conduct the analysis,
and make sure to disconnect the database when you are finished.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Connect to database
connect <- DBI::dbConnect(RSQLite::SQLite(), "./data/pbp_db/pbp_db")

# Load pbp_db
pbp_db <- dplyr::tbl(connect, "nflfastR_pbp")

# Additional code here...

# Close the database when finished
DBI::dbDisconnect(connect)
```

## Calling Functions with Database

Now that a database has been created and saved, `nuclearff` functions can be
utilized.
For example, try obtaining advanced RB stats for a season but using the function
`get_rb_adv_stats_season()`.
The default arguments is to look for a play-by-play database stored in
`data/pbp_db` named `pbp_db` with a table called `nflfastR_pbp`.
If the user database is saved elsewhere or with different names, then define
the argument with a string.
After loading the required packages, the function can be called.
```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(DBI)
library(RSQLite)
library(dplyr)
library(nflfastR)
library(stringr)
library(nuclearff)

rb_adv_stats <- get_rb_adv_stats_season(pbp_db = "./data/pbp_db/pbp_db",
                                        pbp_db_tbl = "nflfastR_pbp",
                                        season = 2024)
```

## Additional Information

To determine if any data from `nflverse` has been updated, check the
[nflverse Data Update and Availability Schedule](https://nflreadr.nflverse.com/articles/nflverse_data_schedule.html).

For more information, refer to the `nflfastR` package [@nflfastR] or the
[nflfastR documentation](https://www.nflfastr.com).

## References
